---
title: 'Homes Impacted by the Houston, TX Blackout (2021)'
description: "Throughout the month of February in 2021, the state of Texas suffered a major power crisis, which came about as a result of three severe winter storms sweeping across the United States on February 10–11, 13–17, and 15–20.In particular, we will use the VNP46A1 to detect differences in night lights before and after the storm to identify areas that lost electric power. To determine the number of homes that lost power, we will link (spatially join) these areas with OpenStreetMap data on buildings and roads. By linking our analysis with data from the US Census Bureau we can investigate the potential socioeconomic factors that influenced recovery."
author: "Sofia Ingersoll"
date: "2023-12-14"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
categories: [Geospatial Analysis, Power Outage]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Throughout the month of February in 2021, the state of Texas suffered a major power crisis, which came about as a result of three severe winter storms sweeping across the United States on February 10--11, 13--17, and 15--20."[^1] For more background, check out these [engineering](https://www.youtube.com/watch?v=08mwXICY4JM&ab_channel=PracticalEngineering) and [political](https://www.youtube.com/watch?v=Zcrsgdl_hP0&ab_channel=Vox) perspectives. For this project, we will estimate the number of homes in Houston that lost power as a result of the first two storms and investigate if socioeconomic factors are predictors of communities recovery from a power outage. Our analysis will be based on remotely-sensed night lights data, acquired from the [Visible Infrared Imaging Radiometer Suite (VIIRS)](https://en.wikipedia.org/wiki/Visible_Infrared_Imaging_Radiometer_Suite) onboard the Suomi satellite. In particular, we will use the VNP46A1 to detect differences in night lights before and after the storm to identify areas that lost electric power. To determine the number of homes that lost power, we will link (spatially join) these areas with [OpenStreetMap](https://www.openstreetmap.org/#map=4/38.01/-95.84) data on buildings and roads. By linking our analysis with data from the US Census Bureau we can investigate the potential socioeconomic factors that influenced recovery.

[^1]: Wikipedia. 2021. "2021 Texas power crisis." Last modified October 2, 2021. <https://en.wikipedia.org/wiki/2021_Texas_power_crisis>.

##### Learning objectives:

-   load vector/raster data\
-   simple raster operations\
-   simple vector operations\
-   spatial joins

## Libraries

```{r message = FALSE}
# Reading in Libraries
library(tidyverse)
library(sf)
library(raster)
library(tmap)
library(terra)
library(stars)
library(patchwork)
```

# Finding Locations of Blackouts

### Night Lights Data

Use NASA's Worldview to explore the data around the day of the storm. There are several days with too much cloud cover to be useful, but 2021-02-07 and 2021-02-16 provide two clear, contrasting images to visualize the extent of the power outage in Texas.

VIIRS data is distributed through NASA's [Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC)](https://ladsweb.modaps.eosdis.nasa.gov/). Many NASA Earth data products are distributed in 10x10 degree tiles in sinusoidal equal-area projection. Tiles are identified by their horizontal and vertical position in the grid. Houston lies on the border of tiles h08v05 and h08v06. We therefore need to download two tiles per date.

-   `VNP46A1.A2021038.h08v05.001.2021039064328.h5.tif`: tile h08v05, collected on 2021-02-07
-   `VNP46A1.A2021038.h08v06.001.2021039064329.h5.tif`: tile h08v06, collected on 2021-02-07
-   `VNP46A1.A2021047.h08v05.001.2021048091106.h5.tif`: tile h08v05, collected on 2021-02-16
-   `VNP46A1.A2021047.h08v06.001.2021048091105.h5.tif`: tile h08v06, collected on 2021-02-16

## Reading VIIRS Data

Reading in night lights tiles using `stars`:

```{r echo=FALSE}
# Reading in the night light tif files
VIIRS_05_07 <- read_stars("../data/houston/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif")
VIIRS_06_07 <- read_stars("../data/houston/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif")
VIIRS_05_16 <- read_stars("../data/houston/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif")
VIIRS_06_16 <- read_stars("../data/houston/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif")
```

## Combining Tiles

Combining tiles into a single `stars` object for each date (2021-02-07 and 2021-02-16).

```{r include=TRUE}
# Combining tiles for each date as a stars object
feb_07 <- st_mosaic(VIIRS_05_07, VIIRS_06_07)
feb_16 <- st_mosaic(VIIRS_05_16, VIIRS_06_16)

```

# Making a Mask

To find the change in night-light intensity (presumably) caused by the storm, we must make a mask to conceal the data that experienced a drop less than 200 nW cm^-2^sr^-1^. This reclassification of the difference raster was done assuming that any location that experienced a drop of more than 200 nW cm^-2^sr^-1^ experienced a blackout. All **locations not considered blackout zones** were assigned `NA.`

```{r include=TRUE}
# Creating a mask to identify locations that experienced a drop greater than 200 nW cm-2sr-1
blackout_locations <- (feb_07 - feb_16) > 200

# Assigning non-blackout locations to be NA values
blackout_locations[blackout_locations == FALSE] <- NA
```

# Vectorized Mask

To vectorize the blackout mask, apply `st_as_sf()` and correct invalid geometries by piping `st_make_valid`.

```{r include=TRUE}
# Transforming the mask into a vector and fixing improper geometries
vector_blackout_locations <- st_as_sf(blackout_locations) %>% 
  st_make_valid(vector_blackout_locations)

# Displays CRS of vector data
#glimpse(crs(vector_blackout_locations))
```

# Cropping a Vectorized Map for Region of Interest

The Houston metropolitan area was initially defined using the following coordinates:

`(-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29)`

These coordinates were then turned into a polygon using `st_polygon` and further converted into a simple feature collection (sfc) using `st_sfc()`. The CRS of this sfc was then converted to match the map to perform the crop.

## Special Feature Collection

```{r include=TRUE}
# Defining the Region of Interest
houston <- st_polygon(list(rbind(c(-96.5, 29),
                                 c(-96.5, 30.5),
                                 c(-94.5, 30.5),
                                 c(-96.5, 29))))

# Creating a sfc and assigning the CRS to match the vector blackout data
houston <- st_sfc(houston) %>% 
  st_set_crs(4326) %>% 
  st_make_valid()

# Confirms last step was successful
#glimpse(crs(houston))
```

## Making a Mask

A cropped version of the blackout map was created using `st_crop` , this mask of the region of interest was then re-projected to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area).

```{r}
# cropping blackout locations to the houston dimensions
houston_blackout_mask <- st_crop(vector_blackout_locations, houston)

# Re-projecting to match the maps
houston_blackout_mask <- houston_blackout_mask %>% 
  st_transform(3083) %>% 
  st_make_valid(houston_blackout_mask)
```

```{r echo = FALSE, include = TRUE}
plot(houston_blackout_mask)
```

# Excluding Highways from the Blackout Map

### Roads Data

Typically highways account for a large portion of the night lights observable from space (see Google's [Earth at Night](https://earth.google.com/web/@27.44405464,-84.7693044,206.63660162a,8916361.52264659d,35y,0h,0t,0r/data=CiQSIhIgMGY3ZTJkYzdlOGExMTFlNjk5MGQ2ZjgxOGQ2OWE2ZTc)). To minimize falsely identifying areas with reduced traffic as areas without power, we will ignore areas near highways.

[OpenStreetMap (OSM)](https://planet.openstreetmap.org/) is a collaborative project which creates publicly available geographic data of the world. Ingesting this data into a database where it can be subsetted and processed is a large undertaking. Fortunately, third party companies redistribute OSM data. We used [Geofabrik's download sites](https://download.geofabrik.de/) to retrieve a shapefile of all highways in Texas and prepared a Geopackage (`.gpkg` file) containing just the subset of roads that intersect the Houston metropolitan area. 

-   `gis_osm_roads_free_1.gpkg`

### Reading in Highway Data

Using `sf` package and `st_read()` to load in the highway data. Additionally, SQL, a language used for storing, manipulating, and retrieving data, was employed to optimize our workflow. A SQL query is describes the mechanism in which the data will be read in. The SQL query is stored as query below. The data was then re-projected to match the CRS of the map data EPSG:3083

```{r echo=FALSE}
# SQL query to aid in the data loading process
query <- "SELECT* FROM gis_osm_roads_free_1 WHERE fclass='motorway'"

# Loading in the data with the SQL query
highways <- st_read("../data/houston/gis_osm_roads_free_1.gpkg",
                    query = query,
                    quiet = TRUE)

# Re-projecting CRS
highways <- highways %>% 
  st_transform(3083) %>% 
  st_make_valid(highways)
```

### Creating a Buffer

Using `st_buffer` to create an undissolved area within 200 m of all highways and `st_union` to dissolve the buffers.

```{r include=TRUE}
# Creating a 200 m buffer and dissolving it as a single object
highway_buffer <- st_buffer(highways, dist = 200) 

# combines the geometries within the buffer as one geometric polygon
highway_buffer <- st_union(highway_buffer)

```

A preliminary check to see if the buffer was successful.

```{r echo = FALSE, include = TRUE}
plot(highway_buffer)
```

Now to combine the cropped vector blackout data of Houston with the highway buffer we created above using `st_difference`.

We expect a few observations to be removed, but to be relatively the same length.

```{r echo=FALSE, include = TRUE}
# identifying homes boardering the buffer and beyond the buffer
houston_blackout_highways <- st_difference(houston_blackout_mask, highway_buffer)

plot(houston_blackout_highways)
```

# Finding Homes Impacted by Blackouts

### Buildings Data

We can also obtain building data from OpenStreetMap. We downloaded from Geofabrick and prepared a GeoPackage containing only houses in the Houston metropolitan area.

-   `gis_osm_buildings_a_free_1.gpkg`

```{r echo = FALSE}
# SQL Query to aid in data loading process
building_query <- "SELECT*
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"

# Loading in the building data with the query
buildings <- st_read("../data/houston/gis_osm_buildings_a_free_1.gpkg",
                     query = building_query,
                     quiet = TRUE)
```

```{r echo=FALSE}
# Re-projecting the dataset to match map
buildings <- buildings %>%
  st_transform('EPSG:3083')
```

## Quantifying Homes Within Blackout Areas

Using the `st_intersection` function to filter for homes beyond the buffer that were effected by blackouts to estimate the number of homes effected by the blackout to be 168,873.

```{r include=TRUE, echo=FALSE}
# Filtering for homes beyond the buffer zone affected by the blackouts
houston_blackout_homes <- st_intersection(buildings, houston_blackout_highways)
# the number of homes impacted
nrow(houston_blackout_homes)
#Need to convert st_intersection to st_intersects
```

Preliminary visualization of Houston Homes effected by the blackout

```{r echo=FALSE}
plot(houston_blackout_homes$geom)
```

# Investigating Socioeconomic Factors

### Socioeconomic Data

We cannot readily get socioeconomic information for every home, so instead we obtained data from the [U.S. Census Bureau's American Community Survey](https://www.census.gov/programs-surveys/acs) for census tracts in 2019. The *folder* `ACS_2019_5YR_TRACT_48.gdb` is an ArcGIS ["file geodatabase"](https://desktop.arcgis.com/en/arcmap/latest/manage-data/administer-file-gdbs/file-geodatabases.htm), a multi-file proprietary format that's roughly analogous to a GeoPackage file.

The geodatabase contains a layer holding the geometry information, separate from the layers holding the ACS attributes. You have to combine the geometry with the attributes to get a feature layer that `sf` can use.

### Read in Socioeconomic Data with sf

Using `st_read()` to load the geodatabase layers and `st_layers()` to explore the contents of the geodatabase. Each layer contains a subset of the fields documents in the [ACS metadata](https://www2.census.gov/geo/docs/maps-data/data/tiger/prejoined/ACSMetadata2011.txt). Geometries are stored in the `ACS_2019_5YR_TRACT_48_TEXAS` layer, income data is stored in the `X19_INCOME` layer, and the median income field `B19013e1.`

```{r echo=FALSE}
# Loading geometry data
socioeconomic <- st_read("../data/houston/ACS_2019_5YR_TRACT_48_TEXAS.gdb", layer = "ACS_2019_5YR_TRACT_48_TEXAS",
                         quiet = TRUE)  

# Loading income data and renaming the columns
income <- st_read("../data/houston/ACS_2019_5YR_TRACT_48_TEXAS.gdb", layer = "X19_INCOME",
                  quiet = TRUE) %>%
  dplyr::select(GEOID, B19013e1) %>% 
  rename(GEOID_Data = GEOID,
         median_income = B19013e1)
```

```{r include=TRUE, echo=FALSE}
# Combining the data to region of interest
census_income <- left_join(socioeconomic, income, by = "GEOID_Data") %>%
  st_transform('EPSG:3083') 

# Re=projecting CRS
houston <- houston %>% 
  st_transform(3083)

# Cropping the data 
houston_income <- st_crop(census_income, houston)

# Census tracts that experienced blackout
blackout_census <- houston_blackout_homes[houston_income,] %>% 
  mutate(blackout = 'yes')

# Combining Cropped datums
houston_blackout_combined <- st_join(houston_income, blackout_census, left = TRUE)

# Additional check to track impact of blackout
houston_blackout_combined$VNP46A1.A2021038.h08v05.001.2021039064328.tif[is.na(houston_blackout_combined$VNP46A1.A2021038.h08v05.001.2021039064328.tif)] <- "FALSE"

# labels unaffected areas
houston_blackout_combined$blackout[is.na(houston_blackout_combined$blackout)] <- "no"

# isolates unaffected areas into data subset
no_blackout <- st_as_sf(houston_blackout_combined[!grepl("yes",houston_blackout_combined$blackout),])
```

```{r echo=FALSE}
# Tracking blackouts
trac_with_blackout <- houston_blackout_combined%>%
  filter(houston_blackout_combined$VNP46A1.A2021038.h08v05.001.2021039064328.tif == TRUE)

# Tracking no blackouts
trac_without_blackout <- houston_blackout_combined%>%
  filter(houston_blackout_combined$VNP46A1.A2021038.h08v05.001.2021039064328.tif == FALSE)
```

## Figure 1: Census Tracts Median Income Examining Blackout Impacts

-   A map of median income by census tract, designating which tracts had blackouts with bubbles.

```{r}
tmap_mode('plot')

#Create a map of median income by census tract, designating which tracts had blackouts
tm_shape(houston_income) + 
  tm_polygons(fill = "median_income", 
          title = "Median Income",
          palette = "magma") +
  tm_compass( type = 'rose', 
              lw = 0.,
              size = 3) +
  tm_bubbles(size = 0.6,
            col = 'black',
             alpha = 0.5,
             fill_alpha.scale = 'black') +
    tm_scale_bar(position = c('right','bottom')) +
  tm_title(text = "Median Census Tract Income in Houston Affected by Blackouts", size = 1) 

```

## Figure 2: Median Income & Examining Blackout vs No Blackouts

-   A map of median income by census tract, designating which tracts had blackouts with bubbles.

```{r}
#census tracts with blackouts
blackout_map <-  tm_shape(trac_with_blackout) + 
  tm_polygons(fill = "median_income", 
          title = "Median Income ($)",
          palette = "magma") +
  tm_borders() +
  tm_title(text = "Median Income for Census Tracts With Blackouts") +
   tm_scalebar(position = c('right','bottom')) + 
  tm_compass( type = 'rose', 
              lw = 0.,
              size = 2) 
#census tracts with no blackouts   
no_blackout_map <-  tm_shape(no_blackout) + 
  tm_polygons(fill = "median_income", 
          title = "Median Income ($)",
          palette = "magma") +
  tm_borders() +
  tm_title(text = "Median Income for Census Tracts Without Blackouts") +
   tm_scalebar(position = c('right','bottom')) + 
  tm_compass( type = 'rose', 
              lw = 0.,
              size = 2) 
  
tmap_arrange(blackout_map, no_blackout_map)

```

## Figure 3: Histogram: Median Income Distribution & Examining Blackout vs No Blackouts

-   Histogram of the distribution of income in impacted and unimpacted tracts.

```{r echo=FALSE, message=FALSE}
#side by side histograms of median income and whether or not they experienced a blackout.
par(mfrow=c(1,2))
#histogram of the distribution of income in impacted tracts
median_income_blackout <- (houston_income$median_income)
hist(median_income_blackout,
main = "Median Income for Tracts Impacted by Blackout",
cex.main = .75,
xlab = "Median Income ($)",
ylab = "Number of Homes",
col = "hotpink",
n = 30)
#histogram of the distribution of income in unimpacted tracts
median_income_no_blackout <- (no_blackout$median_income)
hist(median_income_no_blackout,
main = "Median Income for Tracts Not Impacted by Blackout",
cex.main = .75,
xlab = "Median Income ($)",
ylab = "Number of Homes",
col = "pink",
n = 30)


hist <-  median_income_blackout + median_income_no_blackout 
hist
```

## Figure 4: Boxplot Median Income Distribution & Examining Blackout vs No Blackouts

-   Boxplot of the distribution of income in impacted and unimpacted tracts.

```{r}
ggplot(data = houston_blackout_combined, aes(x = blackout,
                                             y = median_income,
                                             col = blackout)) +
  geom_jitter(width = 0.1) +
  geom_boxplot(alpha = 0.5) +
  labs(x = "Blackout", y = "Median Income",
       title = "Houston Power Outage Experience Considering Median Income Distribution") +
  scale_color_brewer(palette = "Accent")
```

## Conclusion

During the 2021 February string of storms in Houston, a number of homes were effected by power outages.

Approximately 168,956 homes were impacted by the blackout that occurred on February 16, 2023. As a result, a variety of individuals were affected by the blackout. An inspection regarding the relationship between Median Income collected from census data and experiencing a blackout. Only 373 homes were unaffected by the blackouts, but a wide distribution of median wealth were inside this categorization. As expected, Houston homes in census tract regions with lower median incomes experienced a greater number of blackouts compared to regions with higher median incomes.

This census data is limited to the time the census was recorded and does not contain up-to-date socioeconomic information for every household. Further limitations to this analysis are the visualizations. The data wrangling process utilized operations that altered the geometry of the data. Objects such as buildings were spliced into pieces when combining the data using `st_difference` and `st_intersection`. These steps were unavoidable because other functions like `st_intersects` did not preserve the datums sf sticky geometries and produce a sparse geometry binary predicate list (sgbp). SGBP are not compatible with the type of mapping exercised above and cannot be made into sf data frames.
